Concurrent Programming with Harmony

### 2020-07-06 book/paper.tex
Rename CXL --> Harmony

\newcommand{\cxlsource}[1]{
\begin{tabbing}
XX\=XXX\=XXX\kill
    \input{sources/#1.tex}
\end{tabbing}
}

\newcommand{\cxllink}[1]{%
[\href{https://www.cs.cornell.edu/home/rvr/cxl/#1}{\underline{#1}}]%
}

\newcommand{\cxlref}[2]{%
\href{https://www.cs.cornell.edu/home/rvr/cxl/output/#1}{\underline{#2}}%
}

@@

  description={is an atomic operation on the CXL virtual machine, executed by a process}

@@

  description={is a variable that is stored in the memory of the CXL virtual machine and

@@

    a stack.  CXL uses a combination of a stack machine and a register-based machine}

@@

  description={an assignment of values to variables.  In a CXL virtual machine, this includes

@@

        its state.  CXL distinguishes micro steps (machine instructions) and macro

@@

\title{Concurrent Programming with CXL}
\author{Robbert van Renesse}

@@

This book is intended to help people with understanding and
developing concurrent code.  In particular, it uses a new tool
called CXL that helps with \emph{testing} concurrent code.
The approach is based on \emph{model checking}:
\index{model checking}%
instead of relying
on luck, CXL will run \emph{all possible executions} of a particular
test program.  So even if a bug is unlikely to occur, if the test
\emph{can} expose the bug it \emph{will}.  Importantly, if the bug is
found, the model checker precisely shows how to trigger it in
the smallest number of steps.

@@

So what is CXL?
CXL is a concurrent programming language.  It was designed to teach
the basics of concurrent programming, but it is also useful for
testing new concurrent algorithms or even sequential and distributed
algorithms.  CXL programs are not intended to be ``run'' like programs
in most other programming languages---instead CXL programs are
model checked to test that the program has certain desirable
properties and does not suffer from bugs.

The syntax and semantics of CXL is similar to that of Python.
Python is familiar to many programmers and is easy to learn and
use.  We will assume that the reader is familiar with the basics
of Python programming.  We also will assume that the reader
understands some basics of machine architecture and how programs
are executed.  For example, we assume that the reader is familiar
with the concepts of CPU, memory, register, stack, and machine
instructions.

CXL is heavily influenced by Leslie Lamport's work on
TLA+, TLC, and PlusCal~\cite{Lamport02, Lamport09}.
CXL is designed to have a lower learning curve than those
systems, but is not as powerful.  When you finish this book
and want to learn more, we strongly encourage checking
those out.
Another excellent resource is Fred Schneider's book ``On
Concurrent Programming''~\cite{Schneider97}.
(This chapter is named after that book.)

@@

\begin{itemize}
\item \autoref{ch:cxlintro} introduces the CXL programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item \autoref{ch:concurrent} illustrates the problem of
concurrent programming through a simple example in which two processes
are concurrently incrementing a counter.
\item \autoref{ch:cxlmachine} presents the
CXL virtual machine to understand the problem
underlying concurrency better.
\item \autoref{ch:critical} introduces the concept of a
\emph{critical section} and presents various flawed implementations
of critical sections to demonstrate that implementing a critical section
is not trivial.
\item \autoref{ch:peterson} introduces Peterson's Algorithm, an
elegant solution to implementating a critical section.
\item \autoref{ch:method} gives some more details on the CXL
language.
\item \autoref{ch:spinlock} introduces ``hardware'' locks
for implemented critical sections.
\item \autoref{ch:synch} presents CXL modules
that supports locks and other synchronization primitives.
\item \autoref{ch:rdwrbusy} and \autoref{ch:rdwr2} present the
abstraction and implementation of reader/writer
locks, which allow multiple processes in a critical section if they are
all only executing read-only accesses.
\autoref{ch:rdwrbusy} also talks about busy-waiting, which is
an undesirable approach to synchronize processes.
\item \autoref{ch:semaphore} introduces \emph{semaphores},
a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item \autoref{ch:bb} and \autoref{ch:sbs} introduce the
\emph{bounded buffer problem} (aka \emph{producer/consumer problem})
and various solutions.
\item \autoref{ch:starvation} talks about \emph{starvation}:
the problem that in some
synchronization approaches processes may not be able to get access to a
resource they need.
\item \autoref{ch:monitors} presents
\emph{monitors} and \emph{condition variables},
another approach to process synchronication.
\item \autoref{ch:deadlock} describes \emph{deadlock}
where a set of processes are indefinitely waiting for one another to
release a resource.
\item \autoref{ch:actor} presents the \emph{actor model}
and \emph{message passing} as an approach to synchronization.
\item \autoref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two processes communicate reliably over an unreliable
network.
\item \autoref{ch:nonblocking} introduces \emph{non-blocking} or
\emph{wait-free} synchronization algorithms,
which prevent processes waiting for one another more than a bounded number of
steps.
\item \autoref{ch:barrier} describes \emph{barrier synchronization},
useful in high-performance computing applications such as parallel simulations.
\end{itemize}

\chapter{Introduction to Programming with CXL}
\label{ch:cxlintro}

Like Python, CXL is an imperative,
dynamically typed, and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Syntax: Every statement in CXL must be terminated by a semicolon
(even \textbf{while} statements and \textbf{def} statements).
In Python semicolons are optional and rarely used.
There is no syntactic significance to indentation in CXL.
Correct indentation in CXL is encouraged but not enforced.
CXL only supports basic operator precedence or associativity.
Use parentheses liberally to remove ambiguity.
\item CXL does not (currently) support floating point, iterators, or I/O;
CXL does support \textbf{for} loops and various comprehensions over sets.
\item Python is object-oriented, supporting classes with methods and
inheritance; CXL has objects but does not support classes.  CXL supports
pointers to objects.
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in CXL.

\begin{figure}
\begin{code}
\cxlsource{triangle}
\end{code}
\caption{\cxllink{code/triangle.cxl} Computing triangle numbers.}
\label{fig:triangle}
\end{figure}

\autoref{fig:triangle} shows a simple example of a CXL program.
(The code for examples in this book can be found in the \texttt{code} folder under
the name listed in the caption of the example.)
The example is sequential and has a method \texttt{triangle} that takes
an integer number as argument.  Each method has a variable called
\textit{result} that eventually contains the result of the
method (there is no \textbf{return} statement in CXL).  The method
also has a variable called \textit{n} containing the value of the
argument.  The \textit{x..y} notation generates a set containing the numbers
from~\textit{x} to~\textit{y} (inclusive).  The last two lines in the program are
the most interesting.
The first assigns to \textit{x} some unspecified value in the range \texttt{0..N}
and the second verifies that $\mathtt{triangle}(x)$ equals $x(x+1)/2$.

``Running'' this CXL program will try all possible executions, which
includes all possible values for $x$.  Try it out (here \texttt{\$}
represents a shell prompt):

\begin{code}
\begin{verbatim}
$ cxl triangle.cxl
#states = 13
no issues found
$
\end{verbatim}
\end{code}

(For this to work, make sure \texttt{cxl} is in your command shell's search path.)
Essentially, the $\texttt{choose}($S$)$
\index{choose operator}%
operator provides the input to the program by selecting some value from the
set~$S$, while the $\textbf{assert}$ statement checks that the output is
correct.  If the program is correct, the output of CXL is the size of the
``state graph'' (13 states in this case).  If not, CXL also
reports what went wrong, typically by displaying a summary of an execution in
which something went wrong.

In CXL, constants have a default value,
but those can be overridden on the command
line using the \texttt{-c} option.
\index{constant}%
For example, if you want to test the code for \texttt{N = 100}, run:
\begin{code}
\begin{verbatim}
$ cxl -c N=100 triangle.cxl
#states = 103
no issues found
$
\end{verbatim}
\end{code}

\section*{Exercises}
\begin{problems}
\item See what happens if, instead of initializing \textit{result} to 0,
you initialize it to 1.  (You do not need to understand the error report at this time.  They will be explained in more detail in \autoref{ch:cxlmachine}.)
\item Write a CXL program that computes squares by repeated adding.  So the program
should compute the square of $x$ by adding $x$ to an initial value of 0 $x$ times.
\end{problems}

@@

Concurrent programming, aka multithreaded programming, involves multiple
processes
\index{process}%
running in parallel while sharing variables.
\autoref{fig:inc} presents a simple example.
The program
initializes two shared variables: an integer \textit{count} and
an array \textit{done} with two booleans.
Method \texttt{incrementer} takes a parameter called \textit{self}.
It increments \textit{count} and sets \textit{done}[\textit{self}] to \texttt{True}.
Method \texttt{main} waits until \textit{done} flags are set to \texttt{True}.
After that, it verifies that the value of \textit{count} equals 2.
Like Python, CXL supports \texttt{assert} statements with two arguments:
the first is a Boolean condition, and the second is a value that is
reported if the condition fails, in this case
the actual value of \textit{count}.
The program spawns three processes.
The first runs \texttt{incrementer(0)}, the second runs
\texttt{incrementer(1)}, and the last runs \texttt{main()}.
Note that although the processes are \emph{spawned} one at a time,
they will execute concurrently.  It is, for example, quite possible
that \texttt{incrementer(1)} finishes before \texttt{incrementer(0)}
even gets going.

\begin{figure}[h]
\begin{code}
\cxlsource{Up}
\end{code}
\caption{\cxllink{code/Up.cxl} Incrementing twice in parallel.}
\label{fig:inc}
\end{figure}

@@

What is going on is that the CXL program is compiled to machine instructions,
\index{machine instruction}%
and it is the machine instructions that are executed by the underlying CXL
machine.  The details of this appear in \autoref{ch:cxlmachine},
but suffice it to
say that the machine has instructions that load values from memory and store
values into memory.  Importantly, it does not have instructions to atomically
increment or decrement values in memory locations.
So to increment a value in memory,
the machine must do at least three machine instructions.  Conceptually:
\begin{enumerate}
\item load the value from the memory location;
\item add 1 to the value;
\item store the value to the memory location.
\end{enumerate}

When running multiple processes, each essentially runs an instantiation of
the machine, and they do so in parallel.  As they execute, their machine
instructions are interleaved
\index{interleaving}%
in unspecified and often random ways.
A program is correct if it works for any interleaving.
In fact, CXL will try all possible interleavings of the processes
executing machine instructions.

@@

The result in this particular interleaving is that \textit{count} ends up
being 1.
This is known as a \emph{race condition}.
\index{race condition}%
When running CXL, it will
report violations of assertions.  It also provides an example
of an interleaving, like the one above, in which an assertion fails.

@@

The following exercises are intended to show you that on the one hand it
is easy to translate CXL programs into Python, but on the other hand,
it is easier to find concurrency bugs using CXL.

\begin{problems}
\item CXL programs can usually be easily translated into Python.  For example,
\autoref{fig:incpy} is a Python version of \autoref{fig:inc}.
\begin{enumerate}
\item Run \autoref{fig:incpy} using Python.  Does the assertion fail?
\item Using a script, run \autoref{fig:incpy} 1000 times.
For example, if you are using the bash shell, you can do the following:
\begin{code}
\begin{verbatim}
for i in {1..1000}
do
    python Up.py
done
\end{verbatim}
\end{code}
How many times does the assertion
fail (if any).
\end{enumerate}
\item \autoref{fig:incmany} is a version of \autoref{fig:incpy} that has each
incrementer thread increment \textit{count} \texttt{N} times.  Run \autoref{fig:incpy}
10 times (using Python).
Report how many times the assertion fails and what the value of \textit{count}
was for each of the failed runs.
Also experiment with lower values of \texttt{N}.
How large does \texttt{N} need to be for assertions to fail?
\item Can you think of a fix to \autoref{fig:inc}?  Try one or two different fixes
and run them through CXL.  Do not worry about having to come up with a correct fix at this
time---the important thing is to develop an understanding of concurrency.
\end{problems}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../python/Up.py}
\end{code}
\caption{\cxllink{python/Up.py} Python implementation of \autoref{fig:inc}.}
\label{fig:incpy}
\end{figure}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../python/UpMany.py}
\end{code}
\caption{\cxllink{python/UpMany.py} Incrementing \texttt{N} times.}
\label{fig:incmany}
\end{figure}

\chapter{The CXL Virtual Machine}
\label{ch:cxlmachine}
\index{CXL Virtual Machine}%

CXL programs are compiled to CXL bytecode
\index{bytecode}%
(machine instructions for a virtual machine),
which in turn is executed by the CXL virtual machine (CVM).
\index{virtual machine}%
\index{CXL Virtual Machine}%
\index{CVM}%
To understand the problem of concurrent computing, it
is important to have a basic understanding of the CVM.

\section*{CXL Values}
\label{ap:cxlvalues}

CXL programs, and indeed the CVM,  manipulate CXL values.
CXL values are recursively defined:
they include booleans (\texttt{False} and \texttt{True}),
integers (but not floating point numbers),
strings (enclosed by double quotes),
sets of CXL values,
and dictionaries
\index{dictionary}%
that map CXL values to other CXL values.
%
Another type of CXL value is the \emph{atom}.
\index{atom}%
It is essentially
just a name.  An atom is denoted using a period followed by the
name.  For example, \texttt{.main} is an atom.

CXL makes extensive use of dictionaries.
A dictionary maps values, known as \emph{keys}, to values.
The CXL dictionary syntax and properties are a little different from Python.
Unlike Python, any CXL value can be a key, including another
dictionary.
CXL dictionaries are written as
$\mathtt{dict}\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
If $d$ is a dictionary, and $k$ is a key, then the
following expression retrieves the CXL value that $k$ maps to in $d$:
\begin{code}
$d$ $k$
\end{code}
This is unfamiliar to Python programers, but in CXL square brackets can be used
in the same way as parentheses, so you can express the same thing in the form
that is familiar to Python programmers:
\begin{code}
$d$[$k$]
\end{code}
However, if $d$ = \texttt{dict\{ .count: 3 \}}, then you can write
can write $d$.\texttt{count} (which has value 3) instead of having to write
$d$\texttt{[.count]} (although both will work).
Thus using atoms, a dictionary can be made to look much like a Python object.
The meaning of $d$ $a$ $b$ $...$ is ((($d$ $a$) $b$) $...$).

Tuples are special forms of dictionaries where the keys are
the indexes into the tuple.  For example, the tuple
\texttt{(5, False)} is the same CXL value as
\texttt{dict\{ 0:5, 1:False \}}.
The empty tuple \texttt{()} is the same value as \texttt{dict\{\}}.
Note that this is different from the empty set, which is \texttt{\{\}}.
As in Python, you can create singleton tuples by including a comma.
For example, \texttt{(1,)}.

Again, square brackets and parentheses work the same in CXL, so
\texttt{[a, b, c]} (which looks like a Python list)
is the same CXL value as \texttt{(a, b, c)} (which looks like a Python tuple),
which in turn is the same CXL value as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
So if $x$ = \texttt{[ False, True ]},
then $x$[0] = \texttt{False} and $x$[1] = \texttt{True}, just like in Python.
However, when creating a singleton list, make sure you include the
comma, as in \texttt{[False,]}.

CXL is not an object-oriented language, so objects don't have
built-in methods.  However, CXL does have some powerful operators to
make up for some of that.
For example, dictionaries have two handy unary operators.
If $d$ is a
dictionary, then \texttt{keys}~$d$ (or equivalently \texttt{keys}($d$)
returns the set of keys and \texttt{len}~$d$ returns the size of
this set.

Appendix~\ref{ap:cxlvalues} provides details on all the values that
CXL currently supports.

\section*{CXL Bytecode}

A CXL program is translated
into \emph{bytecode}, which is list of machine instructions that the
CXL virtual machine (CVM) executes.
The CVM is not an ordinary virtual machine, but its architecture
is nonetheless representative of conventional computers and
virtual machines such as the Java Virtual Machine.

Instead of bits and bytes, a CVM manipulates CXL values.
A CVM has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of CVM instructions,
generated from a CXL program.  The types of instructions will be described later.
\item Shared memory: A CVM has just one memory location containing
an CXL value.
\item Processes:  Any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable (but not necessarily unique) \emph{name tag},
\index{name tag}%
a program counter,
\index{program counter}%
a stack of CXL values,
and a single mutable general purpose \emph{register}
\index{register}%
that contains a CXL value.
\end{itemize}

@@

The state of a process is called a \emph{context} (aka \emph{continuation}):
\index{context}%
\index{continuation}%
it contains the value of
its name tag, program counter, stack, and register.
The state of a CVM
consists of the value of its memory and the multiset (or \emph{bag})
\index{bag}%
\index{multiset}%
of contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty dictionary, \texttt{dict\{\}}.
The context bag has an initial context in it with name tag
\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
% \autoref{fig:incstate} shows an example of a reachable state for the program in \autoref{fig:inc}.

>>>>

\newcommand{\harmonysource}[1]{
\begin{tabbing}
XX\=XXX\=XXX\kill
    \input{sources/#1.tex}
\end{tabbing}
}

\newcommand{\harmonylink}[1]{%
[\href{https://www.cs.cornell.edu/home/rvr/harmony/#1}{\underline{#1}}]%
}

\newcommand{\harmonyref}[2]{%
\href{https://www.cs.cornell.edu/home/rvr/harmony/output/#1}{\underline{#2}}%
}

@@

  description={is an atomic operation on the Harmony virtual machine, executed by a process}

@@

  description={is a variable that is stored in the memory of the Harmony virtual machine and

@@

    a stack.  Harmony uses a combination of a stack machine and a register-based machine}

@@

  description={an assignment of values to variables.  In a Harmony virtual machine, this includes

@@

        its state.  Harmony distinguishes micro steps (machine instructions) and macro

@@

\title{Concurrent Programming with Harmony}
\author{Robbert van Renesse}

@@

This book is intended to help people with understanding and
developing concurrent code.  In particular, it uses a new tool
called Harmony that helps with \emph{testing} concurrent code.
The approach is based on \emph{model checking}:
\index{model checking}%
instead of relying
on luck, Harmony will run \emph{all possible executions} of a particular
test program.  So even if a bug is unlikely to occur, if the test
\emph{can} expose the bug it \emph{will}.  Importantly, if the bug is
found, the model checker precisely shows how to trigger it in
the smallest number of steps.

@@

So what is Harmony?
Harmony is a concurrent programming language.  It was designed to teach
the basics of concurrent programming, but it is also useful for
testing new concurrent algorithms or even sequential and distributed
algorithms.  Harmony programs are not intended to be ``run'' like programs
in most other programming languages---instead Harmony programs are
model checked to test that the program has certain desirable
properties and does not suffer from bugs.

The syntax and semantics of Harmony is similar to that of Python.
Python is familiar to many programmers and is easy to learn and
use.  We will assume that the reader is familiar with the basics
of Python programming.  We also will assume that the reader
understands some basics of machine architecture and how programs
are executed.  For example, we assume that the reader is familiar
with the concepts of CPU, memory, register, stack, and machine
instructions.

Harmony is heavily influenced by Leslie Lamport's work on
TLA+, TLC, and PlusCal~\cite{Lamport02, Lamport09}.
Harmony is designed to have a lower learning curve than those
systems, but is not as powerful.  When you finish this book
and want to learn more, we strongly encourage checking
those out.
Another excellent resource is Fred Schneider's book ``On
Concurrent Programming''~\cite{Schneider97}.
(This chapter is named after that book.)

@@

\begin{itemize}
\item \autoref{ch:harmonyintro} introduces the Harmony programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item \autoref{ch:concurrent} illustrates the problem of
concurrent programming through a simple example in which two processes
are concurrently incrementing a counter.
\item \autoref{ch:harmonymachine} presents the
Harmony virtual machine to understand the problem
underlying concurrency better.
\item \autoref{ch:critical} introduces the concept of a
\emph{critical section} and presents various flawed implementations
of critical sections to demonstrate that implementing a critical section
is not trivial.
\item \autoref{ch:peterson} introduces Peterson's Algorithm, an
elegant solution to implementating a critical section.
\item \autoref{ch:method} gives some more details on the Harmony
language.
\item \autoref{ch:spinlock} introduces ``hardware'' locks
for implemented critical sections.
\item \autoref{ch:synch} presents Harmony modules
that supports locks and other synchronization primitives.
\item \autoref{ch:rdwrbusy} and \autoref{ch:rdwr2} present the
abstraction and implementation of reader/writer
locks, which allow multiple processes in a critical section if they are
all only executing read-only accesses.
\autoref{ch:rdwrbusy} also talks about busy-waiting, which is
an undesirable approach to synchronize processes.
\item \autoref{ch:semaphore} introduces \emph{semaphores},
a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item \autoref{ch:bb} and \autoref{ch:sbs} introduce the
\emph{bounded buffer problem} (aka \emph{producer/consumer problem})
and various solutions.
\item \autoref{ch:starvation} talks about \emph{starvation}:
the problem that in some
synchronization approaches processes may not be able to get access to a
resource they need.
\item \autoref{ch:monitors} presents
\emph{monitors} and \emph{condition variables},
another approach to process synchronication.
\item \autoref{ch:deadlock} describes \emph{deadlock}
where a set of processes are indefinitely waiting for one another to
release a resource.
\item \autoref{ch:actor} presents the \emph{actor model}
and \emph{message passing} as an approach to synchronization.
\item \autoref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two processes communicate reliably over an unreliable
network.
\item \autoref{ch:nonblocking} introduces \emph{non-blocking} or
\emph{wait-free} synchronization algorithms,
which prevent processes waiting for one another more than a bounded number of
steps.
\item \autoref{ch:barrier} describes \emph{barrier synchronization},
useful in high-performance computing applications such as parallel simulations.
\end{itemize}

\chapter{Introduction to Programming with Harmony}
\label{ch:harmonyintro}

Like Python, Harmony is an imperative,
dynamically typed, and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Syntax: Every statement in Harmony must be terminated by a semicolon
(even \textbf{while} statements and \textbf{def} statements).
In Python semicolons are optional and rarely used.
There is no syntactic significance to indentation in Harmony.
Correct indentation in Harmony is encouraged but not enforced.
Harmony only supports basic operator precedence or associativity.
Use parentheses liberally to remove ambiguity.
\item Harmony does not (currently) support floating point, iterators, or I/O;
Harmony does support \textbf{for} loops and various comprehensions over sets.
\item Python is object-oriented, supporting classes with methods and
inheritance; Harmony has objects but does not support classes.  Harmony supports
pointers to objects.
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in Harmony.

\begin{figure}
\begin{code}
\harmonysource{triangle}
\end{code}
\caption{\harmonylink{code/triangle.hny} Computing triangle numbers.}
\label{fig:triangle}
\end{figure}

\autoref{fig:triangle} shows a simple example of a Harmony program.
(The code for examples in this book can be found in the \texttt{code} folder under
the name listed in the caption of the example.)
The example is sequential and has a method \texttt{triangle} that takes
an integer number as argument.  Each method has a variable called
\textit{result} that eventually contains the result of the
method (there is no \textbf{return} statement in Harmony).  The method
also has a variable called \textit{n} containing the value of the
argument.  The \textit{x..y} notation generates a set containing the numbers
from~\textit{x} to~\textit{y} (inclusive).  The last two lines in the program are
the most interesting.
The first assigns to \textit{x} some unspecified value in the range \texttt{0..N}
and the second verifies that $\mathtt{triangle}(x)$ equals $x(x+1)/2$.

``Running'' this Harmony program will try all possible executions, which
includes all possible values for $x$.  Try it out (here \texttt{\$}
represents a shell prompt):

\begin{code}
\begin{verbatim}
$ harmony triangle.hny
#states = 13
no issues found
$
\end{verbatim}
\end{code}

(For this to work, make sure \texttt{harmony} is in your command shell's search path.)
Essentially, the $\texttt{choose}($S$)$
\index{choose operator}%
operator provides the input to the program by selecting some value from the
set~$S$, while the $\textbf{assert}$ statement checks that the output is
correct.  If the program is correct, the output of Harmony is the size of the
``state graph'' (13 states in this case).  If not, Harmony also
reports what went wrong, typically by displaying a summary of an execution in
which something went wrong.

In Harmony, constants have a default value,
but those can be overridden on the command
line using the \texttt{-c} option.
\index{constant}%
For example, if you want to test the code for \texttt{N = 100}, run:
\begin{code}
\begin{verbatim}
$ harmony -c N=100 triangle.hny
#states = 103
no issues found
$
\end{verbatim}
\end{code}

\section*{Exercises}
\begin{problems}
\item See what happens if, instead of initializing \textit{result} to 0,
you initialize it to 1.  (You do not need to understand the error report at this time.  They will be explained in more detail in \autoref{ch:harmonymachine}.)
\item Write a Harmony program that computes squares by repeated adding.  So the program
should compute the square of $x$ by adding $x$ to an initial value of 0 $x$ times.
\end{problems}

@@

Concurrent programming, aka multithreaded programming, involves multiple
processes
\index{process}%
running in parallel while sharing variables.
\autoref{fig:inc} presents a simple example.
The program
initializes two shared variables: an integer \textit{count} and
an array \textit{done} with two booleans.
Method \texttt{incrementer} takes a parameter called \textit{self}.
It increments \textit{count} and sets \textit{done}[\textit{self}] to \texttt{True}.
Method \texttt{main} waits until \textit{done} flags are set to \texttt{True}.
After that, it verifies that the value of \textit{count} equals 2.
Like Python, Harmony supports \texttt{assert} statements with two arguments:
the first is a Boolean condition, and the second is a value that is
reported if the condition fails, in this case
the actual value of \textit{count}.
The program spawns three processes.
The first runs \texttt{incrementer(0)}, the second runs
\texttt{incrementer(1)}, and the last runs \texttt{main()}.
Note that although the processes are \emph{spawned} one at a time,
they will execute concurrently.  It is, for example, quite possible
that \texttt{incrementer(1)} finishes before \texttt{incrementer(0)}
even gets going.

\begin{figure}[h]
\begin{code}
\harmonysource{Up}
\end{code}
\caption{\harmonylink{code/Up.hny} Incrementing twice in parallel.}
\label{fig:inc}
\end{figure}

@@

What is going on is that the Harmony program is compiled to machine instructions,
\index{machine instruction}%
and it is the machine instructions that are executed by the underlying Harmony
machine.  The details of this appear in \autoref{ch:harmonymachine},
but suffice it to
say that the machine has instructions that load values from memory and store
values into memory.  Importantly, it does not have instructions to atomically
increment or decrement values in memory locations.
So to increment a value in memory,
the machine must do at least three machine instructions.  Conceptually:
\begin{enumerate}
\item load the value from the memory location;
\item add 1 to the value;
\item store the value to the memory location.
\end{enumerate}

When running multiple processes, each essentially runs an instantiation of
the machine, and they do so in parallel.  As they execute, their machine
instructions are interleaved
\index{interleaving}%
in unspecified and often random ways.
A program is correct if it works for any interleaving.
In fact, Harmony will try all possible interleavings of the processes
executing machine instructions.

@@

The result in this particular interleaving is that \textit{count} ends up
being 1.
This is known as a \emph{race condition}.
\index{race condition}%
When running Harmony, it will
report violations of assertions.  It also provides an example
of an interleaving, like the one above, in which an assertion fails.

@@

The following exercises are intended to show you that on the one hand it
is easy to translate Harmony programs into Python, but on the other hand,
it is easier to find concurrency bugs using Harmony.

\begin{problems}
\item Harmony programs can usually be easily translated into Python.  For example,
\autoref{fig:incpy} is a Python version of \autoref{fig:inc}.
\begin{enumerate}
\item Run \autoref{fig:incpy} using Python.  Does the assertion fail?
\item Using a script, run \autoref{fig:incpy} 1000 times.
For example, if you are using the bash shell, you can do the following:
\begin{code}
\begin{verbatim}
for i in {1..1000}
do
    python Up.py
done
\end{verbatim}
\end{code}
How many times does the assertion
fail (if any).
\end{enumerate}
\item \autoref{fig:incmany} is a version of \autoref{fig:incpy} that has each
incrementer thread increment \textit{count} \texttt{N} times.  Run \autoref{fig:incpy}
10 times (using Python).
Report how many times the assertion fails and what the value of \textit{count}
was for each of the failed runs.
Also experiment with lower values of \texttt{N}.
How large does \texttt{N} need to be for assertions to fail?
\item Can you think of a fix to \autoref{fig:inc}?  Try one or two different fixes
and run them through Harmony.  Do not worry about having to come up with a correct fix at this
time---the important thing is to develop an understanding of concurrency.
\end{problems}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../python/Up.py}
\end{code}
\caption{\harmonylink{python/Up.py} Python implementation of \autoref{fig:inc}.}
\label{fig:incpy}
\end{figure}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{../python/UpMany.py}
\end{code}
\caption{\harmonylink{python/UpMany.py} Incrementing \texttt{N} times.}
\label{fig:incmany}
\end{figure}

\chapter{The Harmony Virtual Machine}
\label{ch:harmonymachine}
\index{Harmony Virtual Machine}%

Harmony programs are compiled to Harmony bytecode
\index{bytecode}%
(machine instructions for a virtual machine),
which in turn is executed by the Harmony virtual machine (HVM).
\index{virtual machine}%
\index{Harmony Virtual Machine}%
\index{HVM}%
To understand the problem of concurrent computing, it
is important to have a basic understanding of the HVM.

\section*{Harmony Values}
\label{ap:harmonyvalues}

Harmony programs, and indeed the HVM,  manipulate Harmony values.
Harmony values are recursively defined:
they include booleans (\texttt{False} and \texttt{True}),
integers (but not floating point numbers),
strings (enclosed by double quotes),
sets of Harmony values,
and dictionaries
\index{dictionary}%
that map Harmony values to other Harmony values.
%
Another type of Harmony value is the \emph{atom}.
\index{atom}%
It is essentially
just a name.  An atom is denoted using a period followed by the
name.  For example, \texttt{.main} is an atom.

Harmony makes extensive use of dictionaries.
A dictionary maps values, known as \emph{keys}, to values.
The Harmony dictionary syntax and properties are a little different from Python.
Unlike Python, any Harmony value can be a key, including another
dictionary.
Harmony dictionaries are written as
$\mathtt{dict}\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
If $d$ is a dictionary, and $k$ is a key, then the
following expression retrieves the Harmony value that $k$ maps to in $d$:
\begin{code}
$d$ $k$
\end{code}
This is unfamiliar to Python programers, but in Harmony square brackets can be used
in the same way as parentheses, so you can express the same thing in the form
that is familiar to Python programmers:
\begin{code}
$d$[$k$]
\end{code}
However, if $d$ = \texttt{dict\{ .count: 3 \}}, then you can write
can write $d$.\texttt{count} (which has value 3) instead of having to write
$d$\texttt{[.count]} (although both will work).
Thus using atoms, a dictionary can be made to look much like a Python object.
The meaning of $d$ $a$ $b$ $...$ is ((($d$ $a$) $b$) $...$).

Tuples are special forms of dictionaries where the keys are
the indexes into the tuple.  For example, the tuple
\texttt{(5, False)} is the same Harmony value as
\texttt{dict\{ 0:5, 1:False \}}.
The empty tuple \texttt{()} is the same value as \texttt{dict\{\}}.
Note that this is different from the empty set, which is \texttt{\{\}}.
As in Python, you can create singleton tuples by including a comma.
For example, \texttt{(1,)}.

Again, square brackets and parentheses work the same in Harmony, so
\texttt{[a, b, c]} (which looks like a Python list)
is the same Harmony value as \texttt{(a, b, c)} (which looks like a Python tuple),
which in turn is the same Harmony value as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
So if $x$ = \texttt{[ False, True ]},
then $x$[0] = \texttt{False} and $x$[1] = \texttt{True}, just like in Python.
However, when creating a singleton list, make sure you include the
comma, as in \texttt{[False,]}.

Harmony is not an object-oriented language, so objects don't have
built-in methods.  However, Harmony does have some powerful operators to
make up for some of that.
For example, dictionaries have two handy unary operators.
If $d$ is a
dictionary, then \texttt{keys}~$d$ (or equivalently \texttt{keys}($d$)
returns the set of keys and \texttt{len}~$d$ returns the size of
this set.

Appendix~\ref{ap:harmonyvalues} provides details on all the values that
Harmony currently supports.

\section*{Harmony Bytecode}

A Harmony program is translated
into \emph{bytecode}, which is list of machine instructions that the
Harmony virtual machine (HVM) executes.
The HVM is not an ordinary virtual machine, but its architecture
is nonetheless representative of conventional computers and
virtual machines such as the Java Virtual Machine.

Instead of bits and bytes, a HVM manipulates Harmony values.
A HVM has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of HVM instructions,
generated from a Harmony program.  The types of instructions will be described later.
\item Shared memory: A HVM has just one memory location containing
an Harmony value.
\item Processes:  Any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable (but not necessarily unique) \emph{name tag},
\index{name tag}%
a program counter,
\index{program counter}%
a stack of Harmony values,
and a single mutable general purpose \emph{register}
\index{register}%
that contains a Harmony value.
\end{itemize}

@@

The state of a process is called a \emph{context} (aka \emph{continuation}):
\index{context}%
\index{continuation}%
it contains the value of
its name tag, program counter, stack, and register.
The state of a HVM
consists of the value of its memory and the multiset (or \emph{bag})
\index{bag}%
\index{multiset}%
of contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the Harmony memory is the empty dictionary, \texttt{dict\{\}}.
The context bag has an initial context in it with name tag
\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
% \autoref{fig:incstate} shows an example of a reachable state for the program in \autoref{fig:inc}.

\begin{figure}
\begin{code}
\begin{verbatim}
Up.hny:1 def incrementer(self):
    0 Jump 34
    1 Frame incrementer(self)
Up.hny:2     count = count + 1;
    2 Load count
    3 Push 1
    4 2-ary +
    5 Store count
Up.hny:3     done[self] = True;
    6 LoadVar self
    7 PushAddress done
    8 Address 2
    9 Push True
   10 Store
   11 Return
\end{verbatim}
\end{code}
\caption{The first part of the HVM bytecode corresponding to \autoref{fig:inc}.}
\label{fig:inccode}
\end{figure}
